Hijacking Simulators with Universal Probabilistic Programming

### 2019-07-24 paper/learning/notes.tex
updated

To capture observed relationships between infection and measured entomological exposure, the susceptibility is then:
\begin{equation}
  S(i,t) = (S_{\infty} + \frac{1 - S_{\infty}}{1 + \frac{E_a(i,t)}{E^{*}}})(S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p(i,t)}{X_{p}^{*}})^{\gamma_p}})
\end{equation}
where $S_{imm}, X^{*}_{p} E^*, \gamma_p, S_\infty$ are latent variables
and $X_p(i,t) = \int^{t}_{-a(i,t)} E_{a}(i,\tau) d\tau$
Fitting this model to data of~\cite{beier1994plasmodium} using their method learns the values $S_{\infty} = 0.049$ and $E^* = 0.032$ 
inoculations/person-night. 
The model for each individual infection $j$ in host $i$ comprises a time series of parasite densities.
  In a naïve host the expectation of the logarithm of the density of the infection at time point $\tau$
in the course of the infection, $\ln(y_0(i,j,\tau))$, is taken from a statistical description of parasite densities
in malaria therapy patients~\cite{maire2006model}. 
In the presence of previous exposure and co-infection, the expected log density for each concurrent infection is: 
\begin{equation}
  E(\ln(y(i,j,\tau))) = D_{y}(i,t)D_{m}(i,t)D_{h}(i,t)\ln(y_0(i,j,\tau)) + \ln(\frac{D_x}{M(i,t)} + 1 - D_x))
\end{equation}
where $M(i,t)$ is the total multiplicity of infection and 
\begin{equation}
  D_{y}(i,t) = \frac{1}{1 + \frac{X_y(i,j,t)}{X^{*}_{y}}}
\end{equation}
where $X_y(i,j,t) = \sum^{t}_{t-a}Y(i,t) - \sum^{t}_{t_{0,j}} y(i,j,\tau)$ (note that a continuous time approximation to this
is given in the original publications~\cite{smith2006mathematical,maire2006model}
and hence measures the cumulative parasite load, and  \textbf{TODO: To finish}
  % \begin{align}
%   \mu_i &= N \\ % pop size
%   % disease dynamics
%   S_{\infty} &\sim \mathcal{N}(\mu_\infty, \sigma^{2}_\infty) \\
%   E_{crit} &\sim \mathcal{N}(\mu_{crit}, \sigma^{2}_{crit}) \\
%   S_{imm} &\sim \mathcal{N}(\mu_{imm}, \sigma^{2}_{imm}) \\
%   X_{p_{crit}} &\sim \mathcal{N}(\mu_{p_{crit}}, \sigma^{2}_{p_{crit}}) \\
%   \gamma_p &\sim \mathcal{N}(\mu_{p}, \sigma^{2}_{p}) \\
%   \rho_i & \sim \mathcal{U}(\mathbb{L}_i , \mathbb{U}_i) \\
%   % integration for parasite densities in different demographics
%   for \hspace{0.1cm} t &= 1:t_{max}: \\
%   E_a &= \frac{E_{{max}_t}}{A_{max}} \times h_{\mu_{SA}}() \\
%   X_p &= E_a + X_p \\
%   S_1 &= S_{\infty} + \frac{(1 - S_\infty) }{1 + \frac{E_a}{E_{crit}}} \\
%   S_2 &= S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p}{X_p_{crit}})^{\gamma_p}}\\
%   %Probability of survival function
%   S_p &= S_1 \times S_2 \\
%   \lambda &= S_p \times E_a \\
%   % Number of infections generated in unit time per individual sampled from poisson
%   h &\sim Pos(\lambda) \\
%   \texttt{if} &\hspace{0.1cm} h > 0: \\
%   \hspace{2.3cm} &n_{infect} = 1 \\
%   \text{TODO: add prevelance construction below}
%   % % end for - add prevelance construction below
%   % \mathcal{P}rev &= 
% \end{align}

@@

@@

We start by noting that the prior, defined in the context of the outter samples is 
defined as: 
\begin{equation}
  p(x) = 
  \begin{cases}
    \prod^{n_{x}}_{j=1} f_{a_j}(x_j ; \psi_j) if \mathcal{B}(x) = 1 \\
    0
  \end{cases}
\end{equation}
We start with the information projection we construct the variational objective as follows:
\begin{align}
  J(\eta) = \mathbb{E}_{p(x)}[KL(p(x) || q(x; \eta))] \\
   = \int p(x') [\int p(x)\ln\left(\frac{p(x)}{q(x;\eta)}\right) dx]dx' \\
  %  \int p(x')dx' \gets 1 \\
  %  &= \mathbb{E}_{p(x)}[\ln(p(x))]  - \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\
  %  &= \ln(1) -  \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\ % I do not agree with this line
  %  &= -\mathbb{E}_{p(x)}[\ln(q(x;\eta))] \\
  %  &= -\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1}\ln(q_{a_j}(x_j;\eta))] \\
   \text{Let } a_{j} \in \{1, \ldots, L\} , S_{R} \subset \{1, \ldots, L\} \\
   = \cancelto{\text{no dependence on } \eta}{\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \notin S_{R}}\ln(p(x))]}-\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \in S_{R}}\ln(q_{a_j}(x_j;\eta))]\\
\end{align}

>>>>

To capture observed relationships between infection and measured entomological exposure, the susceptibility is then:
\begin{equation}
  S(i,t) = (S_{\infty} + \frac{1 - S_{\infty}}{1 + \frac{E_a(i,t)}{E^{*}}})(S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p(i,t)}{X_{p}^{*}})^{\gamma_p}})
\end{equation}
where $S_{imm}, X^{*}_{p} E^*, \gamma_p, S_\infty$ are latent variables
and $X_p(i,t) = \int^{t}_{-a(i,t)} E_{a}(i,\tau) d\tau$
Fitting this model to data of~\cite{beier1994plasmodium} using their method learns the values $S_{\infty} = 0.049$ and $E^* = 0.032$ 
inoculations/person-night. 
The model for each individual infection $j$ in host $i$ comprises a time series of parasite densities.
  In a naïve host the expectation of the logarithm of the density of the infection at time point $\tau$
in the course of the infection, $\ln(y_0(i,j,\tau))$, is taken from a statistical description of parasite densities
in malaria therapy patients~\cite{maire2006model}. 
In the presence of previous exposure and co-infection, the expected log density for each concurrent infection is: 
\begin{equation}
  E(\ln(y(i,j,\tau))) = D_{y}(i,t)D_{m}(i,t)D_{h}(i,t)\ln(y_0(i,j,\tau)) + \ln(\frac{D_x}{M(i,t)} + 1 - D_x))
\end{equation}
where $M(i,t)$ is the total multiplicity of infection and 
\begin{equation}
  D_{y}(i,t) = \frac{1}{1 + \frac{X_y(i,j,t)}{X^{*}_{y}}}
\end{equation}
where $X_y(i,j,t) = \sum^{t}_{t-a}Y(i,t) - \sum^{t}_{t_{0,j}} y(i,j,\tau)$ (note that a continuous time approximation to this
is given in the original publications~\cite{smith2006mathematical,maire2006model}
and hence measures the cumulative parasite load, where
\begin{equation}
  D_{y}(i,t) = \frac{1}{1 + \frac{X_{h}(i,t)}{X^{*}_h}}
\end{equation}
where $X_h(i,t) = \sum^{t}_{t-a}h(i,\tau) - 1$, the number of inoculations since birth, excluding
the one under consideration, which measures the inocula experienced by the host
up to the time point under consideration. 
\begin{equation}
  D_{m}(i,t) = 1 - \alpha_m \exp(-\frac{0.693a(i,t)}{a^{*}_m})
\end{equation}
which measures the effect of maternal immunity. 
$X^{*}_{y}$, $X^{*}_h$, $D_x$, $a^*_{m}$ and $\alpha_m$ are essentially 
latent variables to be learnt given the field data.

\paragraph{Mass action model of the force of infection}
The base model assumes that susceptibility is a sigmoidal function of the exposure, which conflicts
with the usual mass action assumption for infectious disease transmission.
However, if entomological exposure is over-dispersed, 
infection-exposure relationships similar to those observed in the field 
can be reconciled with mass action principles, thus providing explanation
for the observed non-linear relationship between infection and entomological exposure
and hence a more justifiable representation of the infection process than that used
in the base model~\cite{smith2008estimation}. 
We simulate this by including random variation in the availability of the human host to mosquitoes.
This variation comprises an individual-specific component which is captured by defining
for each simulated human a baseline availability, $A_b(i)$, which is sampled at birth using:
\begin{equation}
  \log(A_b(i)) \sim \mathcal{N}(-\frac{\sigma_b^{2}}{2}, \sigma^{2}_b)
\end{equation}

so that $A_b(i)$ is log-normally distributed with arithmetic mean 1. 
The age adhusted EIR at time $t$, adjusted to the individual is then:
\begin{equation}
  E_{b}(i,t) = E_{max}(t)A_b(t)\frac{A(a(i,t))}{A_{max}}
\end{equation}
Additional log normal variation is introduced at each time step in the simulation 
to make the expected number of entomological inoculations a function
both of the individual and of log-normal noise measured by the
variance parameter, $\sigma^2_n$ :
\begin{equation}
  \ln(E_{a}(i,t)) \sim Normal(\ln(E_{b}(i,t) - \frac{\sigma^2_n, \sigma^2_n}))
\end{equation}
The susceptibility in these models is independent of EIR, i.e.:
\begin{equation}
  S(i,t( = S_{\infty}(S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p(i,t)}{X_{p}^{*}})^{\gamma_p}})
\end{equation}

Three different parameterisations, XML files, of this model were considered in [8] (E0063,R0065 and R0068).
In each case, the overall variability in $E_{a}(i,t)$ was constrained to the same total,
estimated by fitting to the data of~\cite{beier1994plasmodium}.
A probabilistic programming approach to determining these paramteters would
have provided a way to leverage state of the art inference algorithms that 
enable conditioning on arbitrary models. 
The parameters learnt from the original approach taken by the Malaria community 
generated values of $\sigma_b^2 = 0.5, 0.3$ and $0.05$ for each model respectively, corresponding
to $\sigma^2_n = 0.187, 0.567$ and $0.704$. 
\textbf{TODO: To finish - Add prob prog example + plots}

\paragraph{Models of decay of natural immunity}
\textbf{To complete}
  % \begin{align}
%   \mu_i &= N \\ % pop size
%   % disease dynamics
%   S_{\infty} &\sim \mathcal{N}(\mu_\infty, \sigma^{2}_\infty) \\
%   E_{crit} &\sim \mathcal{N}(\mu_{crit}, \sigma^{2}_{crit}) \\
%   S_{imm} &\sim \mathcal{N}(\mu_{imm}, \sigma^{2}_{imm}) \\
%   X_{p_{crit}} &\sim \mathcal{N}(\mu_{p_{crit}}, \sigma^{2}_{p_{crit}}) \\
%   \gamma_p &\sim \mathcal{N}(\mu_{p}, \sigma^{2}_{p}) \\
%   \rho_i & \sim \mathcal{U}(\mathbb{L}_i , \mathbb{U}_i) \\
%   % integration for parasite densities in different demographics
%   for \hspace{0.1cm} t &= 1:t_{max}: \\
%   E_a &= \frac{E_{{max}_t}}{A_{max}} \times h_{\mu_{SA}}() \\
%   X_p &= E_a + X_p \\
%   S_1 &= S_{\infty} + \frac{(1 - S_\infty) }{1 + \frac{E_a}{E_{crit}}} \\
%   S_2 &= S_{imm} + \frac{1 - S_{imm}}{1 + (\frac{X_p}{X_p_{crit}})^{\gamma_p}}\\
%   %Probability of survival function
%   S_p &= S_1 \times S_2 \\
%   \lambda &= S_p \times E_a \\
%   % Number of infections generated in unit time per individual sampled from poisson
%   h &\sim Pos(\lambda) \\
%   \texttt{if} &\hspace{0.1cm} h > 0: \\
%   \hspace{2.3cm} &n_{infect} = 1 \\
%   \text{TODO: add prevelance construction below}
%   % % end for - add prevelance construction below
%   % \mathcal{P}rev &= 
% \end{align}

@@

\section{Density estimation using Normalising flows}
Notes from models that define an explicit,  
tractable density are highly effective, 
because they permit the use of an optimization algorithm directly on the log-likelihood of the training data.
However,  the  family  of  models  that provide  a  tractable  density  is  limited,
with  different  families  having  different disadvantage.
The rejection sampling loops, if proposed correctly, provide a way of generating 
samples from some density, but do not provide a way of constructing the target density (\textbf{Bradley: Ask Tom, is this true?}). 
Within event and population-based simulators, the rejection sampling loops have been designed with care,
to ensure that they are representative to the events that they are simulating. So we can, with confidence,
assume that they are representative of some density. 
However, to utilise powerful inference engines, we must be able to construct the 
full density of the program.

One recent technique for density estimation, in the amortized inference setting,
is normalising flows combined with variation inference~\cite{rezende2015variational}.
Flow-based methods work by leveraging a simple property of differential geometry applied 
to random variables. 
That being the change of variables.
Let $f(\epsilon)$ be a differentiable transform of noise $\epsilon \sim \pi(\epsilon)$, which models 
some process to generate some data $y = f(\epsilon)$. 
Then the probability density of $y$ under the flow is
obtained by a change of variables: 
\begin{equation}
p(y) = \pi(f^{-1}(y)) \left|\det \frac{\partial f^{-1}}{\partial y}\right|
\end{equation}

@@

We start by noting that the prior, defined in the context of the outter samples is 
defined as: 
\begin{equation}
  p(x) = 
  \begin{cases}
    \prod^{n_{x}}_{j=1} f_{a_j}(x_j ; \psi_j) if \mathcal{B}(x) = 1 \\
    0
  \end{cases}
\end{equation}
We start with the information projection we construct the variational objective as follows:
\begin{align*}
  J(\eta) = KL(p(x) || q(x; \eta)) \\
   = \int p(x)\ln\left(\frac{p(x)}{q(x;\eta)}\right) dx \\
  %  \int p(x')dx' \gets 1 \\
  %  &= \mathbb{E}_{p(x)}[\ln(p(x))]  - \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\
  %  &= \ln(1) -  \mathbb{E}_{p(x)}[\ln(q(x;\eta))]\\ % I do not agree with this line
  %  &= -\mathbb{E}_{p(x)}[\ln(q(x;\eta))] \\
  %  &= -\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1}\ln(q_{a_j}(x_j;\eta))] \\
   \text{Let } a_{j} \in \{1, \ldots, L\} , S_{R} \subset \{1, \ldots, L\} \\
   = \cancelto{\text{no dependence on } \eta}{\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \notin S_{R}}\ln(p(x))]}-\mathbb{E}_{p(x)}[\sum^{n_x}_{j=1,a_j \in S_{R}}\ln(q_{a_j}(x_j;\eta))]\\
\end{align*}
